---
title: "COVID Dashboard data exploration"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

This notebook will be used for two functions:

1) To download and assess the quality of the datasets, and to decide how to wrangle and clean them.
2) To explore the cleaned data sets.

The actual cleaning of the data sets will be handled in the `data_cleaning.R` file. 

#__________________________________
# 1. INSTALL & LOAD LIBRARIES #####
#__________________________________

```{r - library check, echo=FALSE, warning=FALSE, message=FALSE}

if (!require(shiny)) install.packages("shiny", dependencies=TRUE)
if (!require(leaflet)) install.packages("leaflet", dependencies=TRUE)
if (!require(ggplot2)) install.packages("ggplot2", dependencies=TRUE) 
if (!require(readr)) install.packages("readr", dependencies=TRUE)
if (!require(dplyr)) install.packages("dplyr", dependencies=TRUE) 
if (!require(naniar)) install.packages("naniar", dependencies=TRUE)
if (!require(tidyverse)) install.packages("tidyverse", dependencies=TRUE)
if (!require(janitor)) install.packages("janitor", dependencies=TRUE)
if (!require(stringr)) install.packages("stringr", dependencies=TRUE)
if (!require(here)) install.packages("here", dependencies=TRUE)
if (!require(broom)) install.packages("broom", dependencies=TRUE)
if (!require(slider)) install.packages("slider", dependencies=TRUE)
if (!require(lubridate)) install.packages("lubridate", dependencies=TRUE)
if (!require(factoextra)) install.packages("factoextra", dependencies=TRUE)
```

```{r - library load, echo=FALSE, warning=FALSE, message=FALSE}
library(shiny)
library(leaflet)
library(ggplot2)
library(readr)
library(dplyr)
library(naniar)
library(tidyverse)
library(janitor)
library(stringr)
library(here)
library(broom)
library(slider)
library(lubridate)
library(factoextra)
```

#________________________________
# 2.DOWNLOAD & LOAD DATA    #####
#________________________________

```{r - check data subdirectory}
# Ensure the `data_raw` directory exists
if (!dir.exists("data_raw")) {
  dir.create("data_raw")
}
```

```{r - define data urls}
# Define the URLs for the CSV files
urls <- list(
  "total_cases_by_health_board" = "https://www.opendata.nhs.scot/dataset/b318bddf-a4dc-4262-971f-0ba329e09b87/resource/7fad90e5-6f19-455b-bc07-694a22f8d5dc/download/total_cases_by_hb_20231004.csv",
  "total_cases_by_local_authority" = "https://www.opendata.nhs.scot/dataset/b318bddf-a4dc-4262-971f-0ba329e09b87/resource/e8454cf0-1152-4bcb-b9da-4343f625dfef/download/total_cases_by_la_20231004.csv",
  "total_cases_by_age_sex" = "https://www.opendata.nhs.scot/dataset/b318bddf-a4dc-4262-971f-0ba329e09b87/resource/19646dce-d830-4ee0-a0a9-fcec79b5ac71/download/total_cases_agesex_20231004.csv", 
  "total_cases_by_deprivation" = "https://www.opendata.nhs.scot/dataset/b318bddf-a4dc-4262-971f-0ba329e09b87/resource/a965ee86-0974-4c93-bbea-e839e27d7085/download/total_cases_simd_20231004.csv",
  "daily_case_trends_by_health_board" = "https://www.opendata.nhs.scot/dataset/b318bddf-a4dc-4262-971f-0ba329e09b87/resource/2dd8534b-0a6f-4744-9253-9565d62f96c2/download/trend_hb_20231004.csv", 
  "daily_case_trends_by_local_authority" = "https://www.opendata.nhs.scot/dataset/b318bddf-a4dc-4262-971f-0ba329e09b87/resource/427f9a25-db22-4014-a3bc-893b68243055/download/trend_ca_20231004.csv",
  "daily_case_trends_by_age_sex" = "https://www.opendata.nhs.scot/dataset/b318bddf-a4dc-4262-971f-0ba329e09b87/resource/9393bd66-5012-4f01-9bc5-e7a10accacf4/download/trend_agesex_20231004.csv", 
  "daily_case_trends_by_deprivation" = "https://www.opendata.nhs.scot/dataset/b318bddf-a4dc-4262-971f-0ba329e09b87/resource/a38a4c21-7c75-4ecd-a511-3f83e0e8f0c3/download/trend_simd_20231004.csv", 
  "tests_by_health_board" = "https://www.opendata.nhs.scot/dataset/b318bddf-a4dc-4262-971f-0ba329e09b87/resource/8da654cd-293b-4286-96a4-b3ece86225f0/download/test_hb_20231004.csv", 
  "tests_by_local_authority" = "https://www.opendata.nhs.scot/dataset/b318bddf-a4dc-4262-971f-0ba329e09b87/resource/3349540e-dc63-4d6d-a78b-00387b9aca50/download/test_ca_20231004.csv", 
  "daily_covid_hospital_admissions" = "https://www.opendata.nhs.scot/dataset/b318bddf-a4dc-4262-971f-0ba329e09b87/resource/bb4d083b-7b92-4722-85a3-a9b58af1f794/download/daily_covid_admissions_20231004.csv", 
  "ethnicity" = "https://www.opendata.nhs.scot/dataset/b318bddf-a4dc-4262-971f-0ba329e09b87/resource/07b3dff7-3254-4d26-946f-6cd9ff276a1d/download/ethnicity_20231004.csv",
  "hospital_locations" = "https://www.opendata.nhs.scot/dataset/cbd1802e-0e04-4282-88eb-d7bdcfb120f0/resource/c698f450-eeed-41a0-88f7-c1e40a568acc/download/hospitals.csv",
    "council_area_2019" = "https://www.opendata.nhs.scot/dataset/9f942fdb-e59e-44f5-b534-d6e17229cc7b/resource/967937c4-8d67-4f39-974f-fd58c4acfda5/download/ca11_ca19.csv",
  "statistical_qualifiers" = "https://www.opendata.nhs.scot/dataset/2b6f00ec-fee3-4828-9303-89f31b436d2a/resource/b80f9af0-b115-4245-b591-fb22775226c4/download/statisticalqualifiers24052019.csv",
  "ca_pop_est" = "https://www.opendata.nhs.scot/dataset/7f010430-6ce1-4813-b25c-f7f335bdc4dc/resource/09ebfefb-33f4-4f6a-8312-2d14e2b02ace/download/ca2019_pop_est_29012025.csv",
  "hscp_pop_est" = "https://www.opendata.nhs.scot/dataset/7f010430-6ce1-4813-b25c-f7f335bdc4dc/resource/c3a393ce-253b-4c75-82dc-06b1bb5638a3/download/hscp2019_pop_est_29012025.csv",
  "hb_pop_est" = "https://www.opendata.nhs.scot/dataset/7f010430-6ce1-4813-b25c-f7f335bdc4dc/resource/27a72cc8-d6d8-430c-8b4f-3109a9ceadb1/download/hb2019_pop_est_29012025.csv",
  "scotland_pop_proj" = "https://www.opendata.nhs.scot/dataset/9e00b589-817e-45e6-b615-46c935bbace0/resource/7a9e74c9-8746-488b-8fba-0fad7c7866ea/download/scotland_pop_proj_05022025.csv",
  "ca_pop_proj" = "https://www.opendata.nhs.scot/dataset/9e00b589-817e-45e6-b615-46c935bbace0/resource/63d8e0b9-32e2-4847-9353-177f8d931c10/download/ca_pop_proj_03072020.csv",
  "hscp_pop_proj" = "https://www.opendata.nhs.scot/dataset/9e00b589-817e-45e6-b615-46c935bbace0/resource/93137613-2f5c-4d8f-b756-0432abb6fbc0/download/hscp_pop_proj_03072020.csv",
  "hb_pop_proj" = "https://www.opendata.nhs.scot/dataset/9e00b589-817e-45e6-b615-46c935bbace0/resource/0876fc67-05e6-4e87-bc30-c4b0756fff04/download/hb_pop_proj_03072020.csv"
  )
```

```{r - download data}
# Download the files if they don't already exist
for (name in names(urls)) {
  file_path <- paste0("data_raw/", name, ".csv")
  if (!file.exists(file_path)) {
    download.file(urls[[name]], file_path)
    message(paste(name, "downloaded successfully."))
  } else {
    message(paste(name, "already exists. Skipping download."))
  }
}
```

```{r - load data, echo=FALSE, warning=FALSE, message=FALSE}
# Load the datasets
total_cases_by_health_board <- read_csv("data_clean/clean_total_cases_by_health_board.csv")
total_cases_by_local_authority <- read_csv("data_raw/total_cases_by_local_authority.csv")
total_cases_by_age_sex <- read_csv("data_clean/clean_total_cases_by_age_sex.csv")
total_cases_by_deprivation <- read_csv("data_clean/clean_total_cases_by_deprivation.csv")
daily_case_trends_by_health_board <- read_csv("data_clean/clean_daily_case_trends_by_health_board.csv")
daily_case_trends_by_local_authority <- read_csv("data_clean/clean_daily_case_trends_by_local_authority.csv")
daily_case_trends_by_age_sex <- read_csv("data_clean/clean_daily_case_trends_by_age_sex.csv")
daily_case_trends_by_deprivation <- read_csv("data_clean/clean_daily_case_trends_by_deprivation.csv")
tests_by_health_board <- read_csv("data_clean/clean_tests_by_health_board.csv")
tests_by_local_authority <- read_csv("data_clean/clean_tests_by_local_authority.csv")
daily_covid_hospital_admissions <- read_csv("data_clean/clean_daily_covid_hospital_admissions.csv")
ethnicity <- read_csv("data_clean/clean_ethnicity.csv")
hospital_location <- read_csv("data_clean/clean_hospital_location.csv")
council_area_2019 <- read_csv("data_clean/clean_council_area_2019.csv")
statistical_qualifiers <- read_csv("data_raw/statistical_qualifiers.csv")
ca_pop_est <- read_csv("data_clean/clean_ca_pop_est.csv")
hscp_pop_est <- read_csv("data_clean/clean_hscp_pop_est.csv")
hb_pop_est <- read_csv("data_clean/clean_hb_pop_est.csv")
scotland_pop_proj <- read_csv("data_clean/clean_scotland_pop_proj.csv")
ca_pop_proj <- read_csv("data_clean/clean_ca_pop_proj.csv")
hscp_pop_proj <- read_csv("data_clean/clean_hscp_pop_proj.csv")
hb_pop_proj <- read_csv("data_clean/clean_hb_pop_proj.csv")
```

#____________________________________
# 3. ASSESS DATASETS - BASICS  #####
#____________________________________
The following code can be copied and pasted and then adapted as I
explore each dataset. These are the fundamental tools I use to explore a
dataset.

## Initial look at dataset
------------------------------------------------------------------------

```{r - data structure}
# Check the structure of the datasets
str(total_cases_by_local_authority)
str(total_cases_by_health_board)
str(total_cases_by_deprivation)
str(total_cases_by_age_sex)
str(tests_by_local_authority)
str(tests_by_health_board)
str(hospital_location)
str(ethnicity)
str(daily_covid_hospital_admissions)
str(daily_case_trends_by_local_authority)
str(daily_case_trends_by_health_board)
str(daily_case_trends_by_deprivation)
str(daily_case_trends_by_age_sex)
str(council_area_2019)
# str(statistical_qualifiers) # This is a dictionary and doesn't require wrangling or cleaning
str(ca_pop_est)
str(hscp_pop_est)
str(hb_pop_est)
str(scotland_pop_proj)
str(ca_pop_proj)
str(hscp_pop_proj)
str(hb_pop_proj)
```
* `total_cases_by_local_authority`: New reported daily and cumulative counts and rates for positive COVID-19 cases, negative tests and deaths at Council area (local authority) level.
  * Date - [Date] - can use lubridate to separate into more manageable variables (year, month, day, period)
  * CA - [chr] - Council Area - gives a code eg "S12000006" , that can be linked to CAName - also found in `hospital_locations`.
  * CAName - [chr] - Council Area Name - gives a name eg "Clackmannashire" , that can be linked to CA`.
  * NewPositive                         : num [1:32] 7 45 31 37 20 7 20 64 21 16 ...
  * AverageDailyPositive7Days           : num [1:32] 1 6.43 4.43 5.29 2.86 ...
  * TotalCases                          : num [1:32] 22409 50863 54624 42488 42986 ...
  * CrudeRatePositive                   : num [1:32] 43691 34300 44921 39377 44749 ...
  * TotalPCROnly                        : num [1:32] 15472 32453 37242 27022 29642 ...
  * AverageDailyPCROnly7Days            : num [1:32] 0.286 3 2.286 1.857 1.429 ...
  * TotalLFDOnly                        : num [1:32] 5179 14311 13588 12340 10919 ...
  * AverageDailyLFDOnly7Days            : num [1:32] 0.714 3.429 2.143 3.429 1.429 ...
  * TotalLFDAndPCR                      : num [1:32] 1758 4099 3794 3126 2425 ...
  * AverageDailyLFDAndPCR7Days          : num [1:32] NA NA NA NA NA NA NA NA NA NA ...
  * FirstInfections                     : num [1:32] 20537 47615 49596 39466 39554 ...
  * AverageDailyFirstInfections7Days    : num [1:32] 0.714 4.429 3.143 2.857 1.571 ...
  * Reinfections                        : num [1:32] 1872 3248 5028 3022 3432 ...
  * AverageDailyReinfections7Days       : num [1:32] 0.286 2 1.286 2.429 1.286 ...
  * PercentReinfections                 : num [1:32] 8.35 6.39 9.2 7.11 7.98 ...
  * AverageDailyPercentReinfections7Days: num [1:32] 28.6 31.1 29 45.9 45 ...
 
* `total_cases_by_health_board`:
  *
* `total_cases_by_deprivation`:
  *
* `total_cases_by_age_sex`:
  *
* `tests_by_local_authority`:
  *
* `tests_by_health_board`:
  *
* `hospital_location`:
  *
* `ethnicity`:
  *
* `daily_covid_hospital_admissions`:
  *
* `daily_case_trends_by_local_authority`:
  *
* `daily_case_trends_by_health_board`:
  *
* `daily_case_trends_by_deprivation`:
  *
* `daily_case_trends_by_age_sex`:
  *

```{r - summary statistics}
# Summary statistics
summary(total_cases_by_local_authority)
summary(total_cases_by_health_board)
summary(total_cases_by_deprivation)
summary(total_cases_by_age_sex)
summary(tests_by_local_authority)
summary(tests_by_health_board)
summary(hospital_location)
summary(ethnicity)
summary(daily_covid_hospital_admissions)
summary(daily_case_trends_by_local_authority)
summary(daily_case_trends_by_health_board)
summary(daily_case_trends_by_deprivation)
summary(daily_case_trends_by_age_sex)
summary(council_area_2019)
# summary(statistical_qualifiers)  # This is a dictionary and doesn't require wrangling or cleaning
summary(ca_pop_est)
summary(hscp_pop_est)
summary(hb_pop_est)
summary(scotland_pop_proj)
summary(ca_pop_proj)
summary(hscp_pop_proj)
summary(hb_pop_proj)
```

```{r - Head rows}
# View the first few rows
head(total_cases_by_local_authority)
head(total_cases_by_health_board)
head(total_cases_by_deprivation)
head(total_cases_by_age_sex)
head(tests_by_local_authority)
head(tests_by_health_board)
head(hospital_location)
head(ethnicity)
head(daily_covid_hospital_admissions)
head(daily_case_trends_by_local_authority)
head(daily_case_trends_by_health_board)
head(daily_case_trends_by_deprivation)
head(daily_case_trends_by_age_sex)
head(council_area_2019)
# head(statistical_qualifiers)  # This is a dictionary and doesn't require wrangling or cleaning
head(ca_pop_est)
head(hscp_pop_est)
head(hb_pop_est)
head(scotland_pop_proj)
head(ca_pop_proj)
head(hscp_pop_proj)
head(hb_pop_proj)
```

## Missing Data

------------------------------------------------------------------------
The majority of the missing data are in the qualifier columns, e.g, `hbqf`, `sex_qf`, etc.
The qualifier column contains NA values to indicate that no data quality issues were identified for those records. Replacing NA values with 0 or another placeholder would obscure this information and potentially misrepresent the data. Retaining NA values ensures transparency and preserves the integrity of the dataset, allowing for accurate interpretation and analysis.

```{r - Missing data plot}
# Visualize missing data
gg_miss_var(total_cases_by_local_authority) #Orkney's `AverageDailyPCROnly7Days` has an NA
gg_miss_var(total_cases_by_health_board)
gg_miss_var(total_cases_by_deprivation)
gg_miss_var(total_cases_by_age_sex)
gg_miss_var(tests_by_local_authority)
gg_miss_var(tests_by_health_board)
gg_miss_var(hospital_location)
gg_miss_var(ethnicity)
gg_miss_var(daily_covid_hospital_admissions)
gg_miss_var(daily_case_trends_by_local_authority)
gg_miss_var(daily_case_trends_by_health_board)
gg_miss_var(daily_case_trends_by_deprivation)
gg_miss_var(daily_case_trends_by_age_sex)
gg_miss_var(council_area_2019)
gg_miss_var(ca_pop_est)
gg_miss_var(hscp_pop_est)
gg_miss_var(hb_pop_est)
gg_miss_var(scotland_pop_proj)
gg_miss_var(ca_pop_proj)
gg_miss_var(hscp_pop_proj)
gg_miss_var(hb_pop_proj) #hb_qf and sex_qf is looking at the qualifiers of the data, and for most rows it is not applicable and so NA is 
```

This code is to be used if it becomes necessary to remove NA values from the datasets.
```{r - replace NA}
# Handle missing values by replacing with 0
# total_cases_by_local_authority <- total_cases_by_local_authority %>%
#   mutate(across(everything(), ~ifelse(is.na(.), 0, .)))
```



#___________________________________________________________
# 4. DATA EXPLORATION - total_cases_by_health_board    #####
#___________________________________________________________

# Check the structure of the dataset str(total_cases_by_health_board)
```{r}
# Summary statistics
summary(total_cases_by_health_board)
```


```{r}
# View the first few rows
head(total_cases_by_health_board)
```


```{r}
# Visualize missing data
gg_miss_var(total_cases_by_health_board)
```



#______________________________________________________________
# 5. DATA EXPLORATION - total_cases_by_local_authority    #####
#______________________________________________________________

# Check the structure of the dataset str(total_cases_by_local_authority)
```{r}
# Summary statistics
summary(total_cases_by_local_authority)
```


```{r}
# View the first few rows
head(total_cases_by_local_authority)
```


```{r}
# Visualize missing data
gg_miss_var(total_cases_by_local_authority)
```
#_______________________________________________________________
# 6. DATA EXPLORATION - total_cases_by_age_sex #####
#_______________________________________________________________

total_cases_by_age_sex <- total_cases_by_age_sex %>% mutate(Date =
as.Date(as.character(Date), format = "%Y%m%d"))

# Convert Date and create new variables for total_cases_by_age_sex

total_cases_by_age_sex <- total_cases_by_age_sex %>% mutate(Date =
ymd(Date), Year = year(Date), Month = month(Date, label = TRUE), Day =
day(Date), Week = week(Date), Weekday = wday(Date, label = TRUE))

#___________________________________________________________________
# 7. DATA EXPLORATION - total_cases_by_deprivation #####
#___________________________________________________________________
total_cases_by_deprivation <- total_cases_by_deprivation %>%
mutate(Date = as.Date(Date, format = "%Y-%m-%d"))

daily_case_trends_by_health_board <- daily_case_trends_by_health_board
%>% mutate(Date = as.Date(Date, format = "%Y-%m-%d"))

#_____________________________________________________________________________
# 8. DATA EXPLORATION - daily_case_trends_by_local_authority #####
#_____________________________________________________________________________
daily_case_trends_by_local_authority <-
daily_case_trends_by_local_authority %>% mutate(Date = as.Date(Date,
format = "%Y-%m-%d"))

# Convert Date and create new variables for daily_case_trends_by_local_authority

daily_case_trends_by_local_authority <-
daily_case_trends_by_local_authority %>% mutate(Date = ymd(Date), Year
= year(Date), Month = month(Date, label = TRUE), Day = day(Date), Week =
week(Date), Weekday = wday(Date, label = TRUE))

#______________________________________________________________________
# 9. DATA EXPLORATION - daily_case_trends_by_age_sex #####
#______________________________________________________________________

daily_case_trends_by_age_sex <- daily_case_trends_by_age_sex %>%
mutate(Date = as.Date(as.character(Date), format = "%Y%m%d"))

# Convert Date and create new variables for daily_case_trends_by_age_sex

daily_case_trends_by_age_sex <- daily_case_trends_by_age_sex %>%
mutate(Date = ymd(Date), Year = year(Date), Month = month(Date, label =
TRUE), Day = day(Date), Week = week(Date), Weekday = wday(Date, label =
TRUE))

#___________________________________________________________________________
# 10. DATA EXPLORATION - daily_case_trends_by_deprivation #####
#___________________________________________________________________________

daily_case_trends_by_deprivation <- daily_case_trends_by_deprivation %>%
mutate(Date = as.Date(Date, format = "%Y-%m-%d"))

# Convert Date and create new variables for daily_case_trends_by_deprivation

daily_case_trends_by_deprivation <- daily_case_trends_by_deprivation
%>% mutate(Date = ymd(Date), Year = year(Date), Month = month(Date,
label = TRUE), Day = day(Date), Week = week(Date), Weekday = wday(Date,
label = TRUE))

#_______________________________________________________________
# 11. DATA EXPLORATION - tests_by_health_board #####
#_______________________________________________________________
tests_by_health_board <- tests_by_health_board %>% mutate(Date =
as.Date(Date, format = "%Y-%m-%d"))

# Convert Date and create new variables for tests_by_health_board

tests_by_health_board <- tests_by_health_board %>% mutate(Date =
ymd(Date), Year = year(Date), Month = month(Date, label = TRUE), Day =
day(Date), Week = week(Date), Weekday = wday(Date, label = TRUE))

#__________________________________________________________________
# 12. DATA EXPLORATION - tests_by_local_authority #####
#__________________________________________________________________

tests_by_local_authority <- tests_by_local_authority %>% mutate(Date =
as.Date(Date, format = "%Y-%m-%d"))

# Convert Date and create new variables for tests_by_local_authority

tests_by_local_authority <- tests_by_local_authority %>% mutate(Date =
ymd(Date), Year = year(Date), Month = month(Date, label = TRUE), Day =
day(Date), Week = week(Date), Weekday = wday(Date, label = TRUE))

#_________________________________________________________________________
# 13. DATA EXPLORATION - daily_covid_hospital_admissions #####
#_________________________________________________________________________

daily_covid_hospital_admissions <- daily_covid_hospital_admissions %>%
mutate(Date = as.Date(Date, format = "%Y-%m-%d"))

daily_covid_hospital_admissions <- daily_covid_hospital_admissions %>%
mutate(Date = ymd(Date), Year = year(Date), Month = month(Date, label =
TRUE), Day = day(Date), Week = week(Date), Weekday = wday(Date, label =
TRUE))

#___________________________________________________
# 14. DATA EXPLORATION - ethnicity #####
#___________________________________________________

ethnicity <- ethnicity %>% mutate(Date = ymd(Date), Year = year(Date),
Month = month(Date, label = TRUE), Day = day(Date), Week = week(Date),
Weekday = wday(Date, label = TRUE))

#___________________________________________________________
# 15. DATA EXPLORATION - hospital_location #####
#___________________________________________________________

```{r}
non_numeric_lng <- !is.numeric(hospital_location$lng)
non_numeric_lat <- !is.numeric(hospital_location$lat)
```
```{r}
blank_lng <- hospital_location$lng == ""
blank_lat <- hospital_location$lat == ""
print(hospital_location[blank_lng | blank_lat, ])
```
```{r}
hospital_location$lng <- trimws(hospital_location$lng)
hospital_location$lat <- trimws(hospital_location$lat)
```


